Quantifying the importance of a tree prior in Bayesian phylogenetics

### 2019-04-04 article.tex
Partially reviewed 'Description'

\verb;pirouette; is written in the R programming language (\cite{R}).
The goal of \verb;pirouette; is to quantify the importance of a tree
prior. It does so by measuring the inference error BEAST2
makes for a given reconstructed phylogeny, 
simulated under the mechanisms of a (usually novel) diversification model, 
which we will refer to as the 'generative tree prior', $\mathit{P}$. 
Such a study will typically construct phylogenies 
for different combinations of the diversification model's parameters, 
to assess under which scenarios the error made by BEAST2 cannot be neglected. 
Although a well-conducted theoretical study uses many replicates, 
this article only shows examples with one replicate, for the sake of clarity.

\verb;pirouette; is very flexible and allows the user 
to specify a big variety of custom settings. 
These settings can be grouped in macro-sections, 
according to how they operate in the pipeline. 
We summarize them in tables~\ref{tab:options} and~\ref{tab:definitions}.

Although many possible tests can be performed, 
we present here what we think are the five most important 
example research questions (ERQ) that \verb;pirouette; can answer. 
For these ERQs we will quantify the importance of a tree prior.

\begin{sidewaystable}
\centering
  \begin{tabular}{|@{}c|p{2.5cm}|p{9cm}|p{4.5cm}@{}|}
    \hline
    \centering
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \textbf{Symbol} & \textbf{Sub-argument} & \textbf{Description} & 
\textbf{Possible values} \\ 
    \hline
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{t_{p}}$ & tree prior & Macroevolutionary speciation model 
      & Yule, BD, CBS, CCP, CEP \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{c_{m}}$ & clock model & Clock for the DNA mutation rates 
      & strict, relaxed log-normal \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{s_{m}}$ & site model & Nucleotide substitution model 
      & JC, HKY, TN93, GTR \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{m_{r}}$ & mutation rate & Pace at which mutation occurs 
      & $m_{r} \in \mathbb{R}_{>0}$\\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{r_{s}}$ & root sequence & DNA sequence at the root of the tree 
      & any combination of a, c, g, t \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{m_{p}}$ & MRCA prior & Prior knowledge of a most recent common 
      ancestor & see documentation \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{m_{s}}$ & MCMC settings & Markov Chain Monte Carlo algorithm setup 
      & see documentation \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{m_{t}}$ & model type & Criterion to select an inference model 
      & Generative, Candidate \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{r_{i}}$ & run if & Condition under which an inference model is 
      used & Always, Best candidate\\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{e_{f}}$ & error function & Specifies how to measure the error 
      & nLTT, $|\gamma|$ \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{\tau_{p}}$ & twin tree prior & Sets the tree prior for the twin 
tree 
      & Yule, BD \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \hline
  \end{tabular}
  \caption{
    Most important parameter options. 
    BD = birth death (\cite{nee1994reconstructed}), 
    CBS = coalescent Bayesian Skyline (\cite{drummond2005bayesian}), 
    CCP = coalescent constant-population, 
    CEP = coalescent exponential-population,
    JC = Jukes-Cantor 1969 (\cite{jukes1969evolution}), 
    HKY = Hasegawa, Kishino and  Yano (\cite{hasegawa1985dating}), 
    TN93 = \cite{tamura1993estimation}, 
    GTR = Generalised time-reversible model (\cite{tavare1986some}).
  }
  \label{tab:options}
\bigskip

@@

The pipeline is summarized by the following steps:
\begin{enumerate}
    \item for a given phylogeny an alignment is simulated under a known 
alignment model;
    \item for this alignment, according to the specified inference conditions, 
an inference model is picked (which may be different from the generative 
alignment model);
    \item the alignment is then used as BEAST2 input to infer a posterior 
distribution of phylogenies;
    \item posterior phylogenies are compared with the given original phylogeny 
to estimate the error made, according to the error measure specified by the 
user;
\end{enumerate}
The pipeline is visualized in Fig.~\ref{fig:pipeline}. 
There is also the option to generate a 'twin tree', 
that goes through the same pipeline. 
The utility of this twin tree will be explained below.

The first step simulates a DNA alignment from a given 
phylogeny (Fig.~\ref{fig:pipeline}, 1a $\rightarrow$ 2a).
This operation is performed according to the alignment model 
specified by the user, which consists in a root DNA sequence, 
a DNA mutation rate and a site model.
A strict clock model is used as the only clock model
type, as the currently used alignment simulation algorithm
only supports a mutation rate that is equal across all the branches.
This step is relatively fast, 
but longer DNA alignments will noticeably slow down the inference step.

The second step (Fig.~\ref{fig:pipeline}, 2a $\rightarrow$ 3a)
starts from experiments as set up by the user. 
We define an experiment as the combination of an inference model 
and the conditions to actually use it in the inference step.
The inference model can be selected in two ways, either using the
generative model or selecting one or more candidates out of a list of candidate 
inference models.
An experiment can use the generative model as the inference model, if the 
generative model is known.
An experiment uses a candidate model if either the generative model is unknown,
or there is an interest in seeing how well non-generative models compete with
the generative model.
Typically, a generative model and/or a best candidate model are used in the 
actual inference.
When using an experiment with a generative model,
it must use the same site and clock model used in the alignment simulation,
as well as the tree prior that underlied the creation of the given phylogeny. 
When using experiments with candidate models,
\verb;pirouette; selects the best candidate based
on the evidence (i.e. marginal likelihood) for the model given the alignment.
The evidence of an inference model is estimated using a nested sampling (NS)
approach, as described in \cite{maturana2018model}. The nested sampling is
performed by \verb;mcbette; [\cite{mcbette}], that calls the 'NS' BEAST2 
package. 
Using BEAST2 packages (in a scripted way) can only be done under Linux and Mac,
therefore the selection of candidate models is not available directly on 
Windows. 
However, even on Windows systems there is the option to perform the model 
selection from browser, using \verb;mcbette;, yet the computation for this 
model selection is
restricted to one hour for free users of the host (Travis CI) continuous
integration service.

>>>>

\verb;pirouette; is written in the R programming language (\cite{R}).
The goal of \verb;pirouette; is to quantify the importance of a tree
prior.
It does so by measuring the inference error
made for a given reconstructed phylogeny, 
simulated under the mechanisms of a (usually novel) diversification model.
We refer to the true prior that generated the given tree
as the 'generative tree prior', $\mathit{P}$.
Many tree priors have a parameter setting in which these are equivalent
to a standard tree prior. For example, a protracted birth-death model
becomes equivalent to a birth-death when speciation rate approaches infinity.
When benchmarking a novel tree prior, 
such a study will typically construct phylogenies 
for different combinations of the diversification model's parameters, 
to assess under which scenarios the inference error cannot be neglected. 
Although a well-conducted theoretical study uses many replicates, 
this article only shows examples with one replicate, for the sake of clarity.

\verb;pirouette; is very flexible and allows the user 
to specify a wide variety of custom settings. 
These settings can be grouped in macro-sections, 
according to how they operate in the pipeline. 
We summarize them in tables~\ref{tab:options} and~\ref{tab:definitions}.

Although many possible tests can be performed, 
we present here what we think are the five most important 
example research questions (ERQ) that \verb;pirouette; can answer. 
These ERQs gradually build up in complexity.
For each of these ERQs we will quantify the importance of a tree prior.

\begin{sidewaystable}
\centering
  \begin{tabular}{|@{}c|p{2.5cm}|p{9cm}|p{4.5cm}@{}|}
    \hline
    \centering
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \textbf{Symbol} & \textbf{Sub-argument} & \textbf{Description} 
      & \textbf{Possible values} \\ 
    \hline
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{t_{p}}$ & tree prior & Macroevolutionary speciation model 
      & Yule [\cite{yule}], BD, CBS, CCP, CEP \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{c_{m}}$ & clock model & Clock for the DNA mutation rates 
      & strict, relaxed log-normal \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{s_{m}}$ & site model & Nucleotide substitution model 
      & JC, HKY, TN93, GTR \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{m_{r}}$ & mutation rate & Pace at which mutation occurs 
      & $m_{r} \in \mathbb{R}_{>0}$\\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{r_{s}}$ & root sequence & DNA sequence at the root of the tree 
      & any combination of a, c, g, t \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{m_{p}}$ & MRCA prior & Prior knowledge of a most recent common 
      ancestor & see documentation \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{m_{s}}$ & MCMC settings & Markov Chain Monte Carlo algorithm setup 
      & see documentation \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{m_{t}}$ & model type & Criterion to select an inference model 
      & Generative, Candidate \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{r_{i}}$ & run if & Condition under which an inference model is 
      used & Always, Best candidate\\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{e_{f}}$ & error function & Specifies how to measure the error 
      & nLTT, $|\gamma|$ \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{\tau_{p}}$ & twin tree prior 
      & Sets the tree prior for the twin tree 
      & Yule, BD \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \hline
  \end{tabular}
  \caption{
    Most important parameter options.
    BD = birth death (\cite{nee1994reconstructed}), 
    CBS = coalescent Bayesian skyline (\cite{drummond2005bayesian}), 
    CCP = coalescent constant-population, 
    CEP = coalescent exponential-population,
    JC = Jukes and Cantor (\cite{jukes1969evolution}), 
    HKY = Hasegawa, Kishino and Yano (\cite{hasegawa1985dating}), 
    TN93 = Tamura and Nei \cite{tamura1993estimation}, 
    GTR = Generalised time-reversible model (\cite{tavare1986some}).
  }
  \label{tab:options}
\bigskip

@@

The pipeline is summarized by the following steps:
\begin{enumerate}
  \item from a given phylogeny an alignment is simulated 
    under a known alignment model;
  \item from this alignment, according to the specified inference conditions, 
    an inference model is picked (which may differ from the 
    generative alignment model);
  \item the alignment and inference model are used 
    to infer a posterior distribution;
  \item the phylogenies in the posterior are compared with the given phylogeny 
    to estimate the error made, according to the error measure specified 
    by the user;
\end{enumerate}
The pipeline is visualized in Fig.~\ref{fig:pipeline}. 
There is also the option to generate a 'twin tree', 
that goes through the same pipeline. 
The utility of this twin tree will be explained below.

The first step simulates a DNA alignment from a given 
phylogeny (Fig.~\ref{fig:pipeline}, 1a $\rightarrow$ 2a)
using the DNA alignment parameters.
The DNA alignment parameters consist of a DNA root sequence, a DNA mutation rate,
a site model and a (strict) clock model.
The DNA root sequence is the DNA sequence the shared common ancestor has,
and is set to four equally-sized mononucleotide blocks by default, as this
helps interpreting the resulting alignment.
In the context of this research, a 'site model' is a nucleotide substitution
model. Supported site models are JC, HKY, TN and GTR. Only the strict
clock model is supported.

The second step (Fig.~\ref{fig:pipeline}, 2a $\rightarrow$ 3a)
selects inference models from a set of experiments. 
We define an experiment $X$ as the combination of an inference model $I$
and the conditions, $C$ to actually use it in the inference step.
There are two main setups of condition $C$: either the attached
inference model $I$ is the generative model and should always be selected,
or the attached inference model is a candidate model, that will be
be selected if it is the best model.
When setting $C$ to be attached to the generative model,
it is implied that attached inference model $I$ uses the 
same site and clock model used in the alignment simulation,
and $I$ uses the same tree prior as used in the creation of the given phylogeny. 
A 'best' model is defined as the inference model with
the highest evidence (a.k.a. marginal likelihood) for that inference model, 
given the alignment simulated in the previous step.
The evidence of an inference model is estimated by nested 
sampling [\cite{maturana2018model}], using the \verb;NS; BEAST2 package.
Scripted use of BEAST2 packages is only possible under under Linux and Mac.
Windows systems can do the model comparison for shorter DNA sequences
using the web interface of \verb;mcbette; [\cite{mcbette}].

\richel{Up until here}
